---
title: "Analysis of neuronal subset"
author: "Maria Puschhof"
date: "2024-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(Seurat)
library(ggplot2)
library(dplyr)
```

In this document, the entire dataset will be subsetted to specfically analyse the neurons. 

## Data import and meta data annotation

```{r read_data}
neuron <- readRDS("../data/out/neurons_all.rds")
```

## Subset the data

Restrict dataset to neurons

```{r select_neurons}
cells_rm <- rownames(neuron@meta.data)[sapply(neuron[["cell_types"]], 
                                              function(z) {z != "neuron"})]
neuron <- subset(neuron, cells = cells_rm, invert=T)
```

Restrict dataset to experimental conditions with rabies virus.

```{r select_rabies}
cells_rm <- rownames(neuron@meta.data)[sapply(neuron[["Sample"]], 
                                              function(z) {z %in% c("ccNeurons", "onlyNeurons")})]
neuron <- subset(neuron, cells = cells_rm, invert=T)
```

Clean up dataset

```{r clean_up}
# Erase module scores
if(length(colnames(neuron@meta.data)) > 21)
  neuron@meta.data[,22:length(colnames(neuron@meta.data))] <- NULL

# Clean up seurat object
neuron[["pca"]] <- NULL
neuron[["integrated.cca"]] <- NULL
neuron[["umap.unintegrated"]] <- NULL
neuron[["umap.integrated"]] <- NULL
neuron[["RNA"]]@layers[["scale.data"]] <- NULL
```

### Data integration

Perform normalization, feature selection, and PCA at per sample basis.

```{r per_sample}
neuron[["RNA"]] <- split(neuron[["RNA"]], f = neuron$plate)
neuron <- FindVariableFeatures(neuron, selection.method = "vst", nfeatures = 4000)
neuron <- ScaleData(neuron)
neuron <- RunPCA(neuron, verbose = FALSE)
neuron <- FindNeighbors(neuron, dims = 1:30, reduction = "pca")
neuron <- FindClusters(neuron, cluster.name = "unintegrated_clusters", 
                       resolution=0.8)
neuron <- RunUMAP(neuron, dims = 1:30, reduction = "pca", 
                  reduction.name = "umap.unintegrated")
```

Integrate the samples using Seurat’s CCA method.

```{r integrate_data}
neuron <- IntegrateLayers(object = neuron, method = CCAIntegration, 
                          orig.reduction = "pca", new.reduction = "integrated.cca",
                          verbose = FALSE)
neuron[["RNA"]] <- JoinLayers(neuron[["RNA"]])
```

### Dimension reduction and clustering

```{r umap}
neuron <- FindNeighbors(neuron, dims = 1:30, reduction = "integrated.cca")
neuron <- FindClusters(neuron, cluster.name = "integrated_clusters", resolution=0.8)
neuron <- RunUMAP(neuron, dims = 1:30, reduction = "integrated.cca", 
                  reduction.name = "umap.integrated")
```

```{r umap_plot}
DimPlot(neuron, reduction = "umap.integrated", group.by = "integrated_clusters")
```

```{r umap_gpf_plot}
DimPlot(neuron, reduction = "umap.integrated", group.by = "gfp_gate", 
        cols = c("lightgrey", "green"))
```

## Cluster composition

```{r gfp_distr}
distr <- table(neuron@meta.data[,which(colnames(neuron@meta.data) == "integrated_clusters")], 
               neuron$gfp_gate)

# Bar graph relative counts
neuron@meta.data$gfp_gate <- factor(neuron@meta.data$gfp_gate, levels = c(TRUE, FALSE))
ggplot(neuron@meta.data, aes(x = integrated_clusters, fill = gfp_gate)) +
      geom_bar(position="fill") +
      scale_fill_manual(values = c('FALSE' = "grey", 'TRUE' = "green")) +
      ylab("Fraction of cells") + xlab("Clusters") + labs(fill = "Connected neurons") +
      theme_minimal()
```

Statistical test for enrichment of connected neurons in individual clusters

```{r gfp_distr_test}

fisher <- fisher.test(distr, simulate.p.value = TRUE, B = 1e5)

# Convert table object to data frame for plotting
distr_plot <- data.frame( 
  "cluster" = rownames(distr),
  "Connected" = distr[,1], 
  "Unconnected" = distr[,2],
  "pct.true" = prop.table(distr, 1)[, "TRUE"]*100
)

# Underlying plot
ggplot(distr_plot, aes(x=cluster, y=pct.true)) +
    geom_bar(stat = "identity", fill = "#1fc600") + 
    theme(panel.background = element_rect(fill = "white")) +
    theme(axis.line = element_line(linewidth = .5, colour = "black")) +
    labs(x = "Cluster", y = "Relative connectivity") +
    annotate("text", x = Inf, y = Inf, hjust = 1.1, vjust = 1.1,
             label = paste("Fisher test p-value:", 
                           format(fisher$p.value, digits = 2)), size = 3)
```

## Saving the data object

```{r save_data}
saveRDS(neuron, file="../data/out/neurons_subset.rds")
```