---
title: "Analysis of entire dataset"
author: "Maria Puschhof"
date: "2024-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(Seurat)
library(ggplot2)
library(dplyr)
```

This document contains the details of processing of the entire dataset.

## Data import and meta data annotation

```{r read_data}
# Import of data and meta data
neuron <- readRDS("../data/RatCells_unfiltered.rds")
meta_sort <- read.csv("../data/sort_data_merged.csv")

# Cut down meta data to relevant cells
meta_sort <- meta_sort[meta_sort$seq_cell_id %in% rownames(neuron@meta.data),]
meta_sort <- meta_sort[match(rownames(neuron@meta.data), meta_sort$seq_cell_id),]

# Assignment of meta data to Seurat object
if(identical(meta_sort$seq_cell_id, rownames(neuron@meta.data))){
  neuron[["cell_id"]] <- meta_sort$cell_id
  neuron[["Sample.Name"]] <- meta_sort$Sample.Name
  neuron[["gfp_gate"]] <- meta_sort$gfp_gate
  neuron[["plate"]] <- meta_sort$plate
}

# Annotation of experimental condition by plate number
plate_to_sample <- c(
  "P3" = "onlyNeurons",
  "P4" = "N2Cneg",
  "P5" = "N2Cpos",
  "P6" = "N2Cpos",
  "P16" = "N2Cneg",
  "P21" = "ccNeurons"
)
neuron[["Sample"]] <- unname(sapply(neuron@meta.data$plate, function(z){
  plate_to_sample[z]
}))
```

## Preprocessing and QC

```{r qc_plots}
# Set QC parameters
nFeature_min <- 5000
nCount_min <- 100000
mito_max <- 4
nFeature_max <- 11250
nCount_max <- 2e6

# Feature count QC
VlnPlot(neuron, features = "nFeature_RNA") + 
  geom_hline(yintercept = nFeature_min, color="red") +
  geom_hline(yintercept = nFeature_max, color="red")

# Total count QC
VlnPlot(neuron, features = "nCount_RNA") + 
  geom_hline(yintercept = nCount_min, color="red") +
  geom_hline(yintercept = nCount_max, color="red")

# Mitochondrial content QC
VlnPlot(neuron, features = "percent.mt") + 
  geom_hline(yintercept = mito_max, color="red")

VlnPlot(neuron, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

Filtering of the data by quality control parameters.

```{r filtering_qc}

neuron <- subset(neuron, subset = nFeature_RNA > nFeature_min & 
                nFeature_RNA < nFeature_max &
                nCount_RNA > nCount_min &
                nCount_RNA < nCount_max &
                percent.mt < mito_max)
```

### Data integration

Perform normalization, feature selection, and PCA at per sample basis.

```{r per_sample}
neuron[["RNA"]] <- split(neuron[["RNA"]], f = neuron$plate)
neuron <- NormalizeData(neuron)
neuron <- FindVariableFeatures(neuron, nfeatures = 4000)
neuron <- ScaleData(neuron)
neuron <- RunPCA(neuron, verbose = FALSE)
neuron <- FindNeighbors(neuron, dims = 1:30, reduction = "pca")
neuron <- FindClusters(neuron, cluster.name = "unintegrated_clusters")
neuron <- RunUMAP(neuron, dims = 1:30, reduction = "pca", 
                  reduction.name = "umap.unintegrated")
```

Integrate the samples using Seurat’s CCA method.

```{r integrate_data}
neuron <- IntegrateLayers(object = neuron, method = CCAIntegration, 
                          orig.reduction = "pca", new.reduction = "integrated.cca",
                          verbose = FALSE)
neuron[["RNA"]] <- JoinLayers(neuron[["RNA"]])
```

### Dimension reduction and clustering

```{r umap}
neuron <- FindNeighbors(neuron, dims = 1:30, reduction = "integrated.cca")
neuron <- FindClusters(neuron, cluster.name = "integrated_clusters")
neuron <- RunUMAP(neuron, dims = 1:30, reduction = "integrated.cca", 
                  reduction.name = "umap.integrated")

```

```{r umap_plot}
DimPlot(neuron, reduction = "umap.integrated", group.by = "integrated_clusters")
```

## Cell state annotation using public references

The cell state classification is based on marker genes of a meta-analysis of GBM studies (Ruiz-Moreno et al., 2022, bioRxiv) as provided by supplementary table 4.

```{r read_gbmap}
gbmap <- read.csv("../data/RuizMoreno_bioRxiv_2022_ST4.csv", skip=1)
# Extract data of level 3 annotation
gbmap <- gbmap[,grep("\\.2$", colnames(gbmap))]
gbmap <- gbmap[gbmap$cluster.2 != "",]
```

Get marker genes for each cell state

```{r marker_genes}
markers <- list()
for(state in unique(gbmap$cluster.2)){
  markers[[state]] <- gbmap$gene.2[gbmap$cluster.2 == state]
}
to_lowercase <- function(strings) {
  strings <- tolower(strings)
  sapply(strings, function(x) paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x))))
}
markers <- lapply(markers, to_lowercase)
```

Calculate similarity to expression profiles in data using Seurat’s module score function

```{r module_score}
for(state in names(markers)){
  neuron <- AddModuleScore(neuron, features=markers[state],
                           name=paste0("int.", state), reduction = "umap.integrated")
}
```

### Identification of astrocytes and oligodendrocytes

```{r score_distribution}
VlnPlot(neuron, features="int.Astrocyte1", group.by="integrated_clusters")

VlnPlot(neuron, features="int.Oligodendrocyte1", group.by="integrated_clusters")
```

Identify clusters with a mean score exceeding the threshold 0.1

```{r cluster_identification}
thres = 0.1
module = "int.Astrocyte1"

# Calculate mean astrocyte score
score_mean <- neuron@meta.data %>% 
  select("integrated_clusters", module) %>%
  group_by(!!sym("integrated_clusters")) %>%
  summarise(mean=mean(!!sym(module)), n=n())

# Identify astrocyte clusters
cl_astro <- as.character(score_mean[score_mean$mean > thres,][["integrated_clusters"]])
print(paste("Astrocyte clusters:", paste(cl_astro, collapse = ", ")))


module = "int.Oligodendrocyte1"

# Calculate mean oligodendrocyte score
score_mean <- neuron@meta.data %>% 
  select("integrated_clusters", module) %>%
  group_by(!!sym("integrated_clusters")) %>%
  summarise(mean=mean(!!sym(module)), n=n())

# Identify oligodendrocyte clusters
cl_oligo <- as.character(score_mean[score_mean$mean > thres,][["integrated_clusters"]])
print(paste("Oligodendrocyte clusters:", paste(cl_oligo, collapse = ", ")))
```

Assign cell types and plot distribution

```{r cell_types}
neuron[["cell_types"]] <- sapply(neuron@meta.data["integrated_clusters"], function(z){
  ifelse(z %in% cl_astro, "astrocyte", ifelse(z %in% cl_oligo, "oligodendrocyte", "neuron"))
})

DimPlot(neuron, reduction = "umap.integrated", group.by = "cell_types",
        cols = c("astrocyte" = "#2a77b5", "oligodendrocyte" = "#e89e20", "neuron" = "#d74749"))
```


## Saving the data object

```{r save_data}
saveRDS(neuron, file="../data/out/neurons_all.rds")
```